Hybrid platforms and bargaining power

### 2023-04-11 src/paper/application.tex
Rewrite introduction

Even though the economics literature offers no unambiguous definition for what a platform is, there seems to be a consensus that they are of key importance for the modern economy, and are getting even more so in the future.
Although market structures resembling various aspects of what we call platforms have existed for some time, the term has only gained widespread usage in the 2000s, with the advent of digitization, and a number of behemoth technology companies playing matchmakers on the internet.
Since then, platforms have become a buzzword in the business world, the subject of increasing regulatory scrutiny, and also a topic of intense academic interest.

In the last couple of decades, a considerable amount of research has been devoted to the study of platforms.
This paper aims to contribute to this literature by examining an important, but somewhat underexplored aspect of platforms: the bargaining power disparity between them and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
In particular, this paper proposes an analytically tractable framework which captures many important aspects of bargaining between one large and a continuum of small players, and applies it to the setting of a hybrid platform facing a continuum of potential entrants.
The results show that when a platform operating in hybrid mode can increase its profits at the cost of decreased seller entry consumer welfare in comparison to the case when the platform is a pure intermediary.
Before an overview of the relevant literature is given, it is useful to briefly discuss the most prominent features of platforms, as that will be useful to understand the relation of this paper to the existing body of research.

While it is hard to define what a platform is, there are a number of features commonly associated with platforms.\footnote{
    I thank the participants of the Third Year Seminar at the University of Zurich for their lively discussion on this topic.
    The conclusion was that, although platforms are hard to define, "You know one when you see one".
}
An organization does not have to exhibit all of these features to be considered a platform, but they are often present, and are also the main areas of interest when studying platforms.
First, platforms are often characterized by multi-sidedness.\footnote{
    The most prominent examples are large internet companies, such as Google connecting advertisers and consumers, Amazon connecting sellers and consumers, or Uber connecting drivers and passengers.
}
Second, a frequent feature of platforms is the presence of network effects.\footnote{
    More buyers on Amazon mean more value for the sellers, and vice versa.
    These network effects often lead to increasing returns to scale, which in turn leads to consolidation in the industry.
}
A third important aspect of platforms is the disparity between the various market participants.
While there is usually just a few, large platforms in an industry, there is a much larger number of participants who use the platforms' services to connect to each other.
Furthermore, the platform is often indispensable for the participants to make business, but the participants themselves are more easily replaceable.
This power imbalance naturally leads to antitrust concerns, with regulators taking an increasingly close look at such markets as of late \parencite[see e.g. ][]{fletcher2021consumer}.
Finally, a somewhat more recent, but slightly alarming feature of platforms is operating in a so-called hybrid mode.
They act as an intermediary (deciding the rules) and a market participant (competing with the other entrants) at the same time.
This paper focuses on the latter two aspects (the interplay between bargaining power and hybrid operation), while mostly abstracting away the first two in the form of a rather reduced-form approach of modeling the production function.

Most of the earlier literature on platform economics focuses on the multi-sidedeness of platforms and the related network effects.
These papers investigate how various market features, such as the possibility of multi-homing, the magnitude of network-effects, or the number of platforms influence market outcomes.
In addition to welfare, they are often interested in the transfers between the various sides facilitated by the platforms' desire to exploit network effects.
Examples of such studies include the seminal papers \textcite{rochet2003platform} and \textcite{armstrong2006competition}, as well as a number of insightful papers examining the effects of various market attributes \parencite[e.g.][]{hagiu2004optimal,choi2010tying,evans2011platform,lee2014competing}.

More recently, hybrid platforms also received increasing attention in the literature.
While not explicitly phrased in terms of hybrid operation, the literature on the importance of exclusive content can be seen as adjacent to this topic \parencite[]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}.
These papers mostly focus on competition between platforms, and not on the impact on the platform-participant relationship.
On the other hand, more recent papers, such as \textcite{hagiu2022should}, \textcite{anderson2021hybrid} and \textcite{gutierrez2021welfare} have started to address the issue of hybrid platforms, and how platforms having their own products affect competition between the platform and the entrants, and ultimately consumer welfare and producer surplus.
The results are mixed: while some of them find that hybrid platforms can be beneficial if certain anti-competitive behavior, such as self-preferencing, is prevented \parencite[]{anderson2021hybrid}, others are more critical of this hybrid form \parencite[]{anderson2021hybrid}.
This paper provides some additional insights into this topic, by looking at the impact of dual mode operation on bargaining power.

Many results and concepts of this paper are also applicable outside the context of (multi-sided) platforms.
Certain structures, such as vertical relationships, or even traditional retail stores share a number of features with platforms (notably, the presence of a dominant, indispensable entity), and can be modeled in much the same way.
Therefore, the results of this paper also contributes to the literature on those settings, such as retailers having their own private labels \parencite{steiner2004nature}, or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
It is particularly closely related to \textcite{de2005vertical} and \textcite{montez2007downstream} in which bargaining between the upstream and downstream firms take the center stage.

Finally, this paper also belongs to the not too large set of models using concepts from cooperative game theory in an industrial organization setting.
Such examples include the aforementioned \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
In contrast to the majority of those, I utilize an oceanic game (a continuum of small players instead of a finite number), and demonstrate that this can considerably simplify the analysis in certain cases.

>>>>

Even though the economics literature offers no unambiguous definition for what platforms are, there seems to be a consensus that they are of key importance for the modern economy, and are getting even more so in the future.
Although market structures resembling various aspects of what we call platforms have existed for some time, the term has only gained widespread usage in the 2000s, with the advent of digitization, and a number of behemoth technology companies playing matchmakers on the internet.
Since then, platforms have become a buzzword in the business world, the subject of increasing regulatory scrutiny, and also a topic of intense academic interest.

Platforms have a number of unique features, which make them different from more traditional market structures.
The most prominent of those are multi-sidedness and network effects, which most of the early literature \parencite[for an overview, see][]{rochet2006two} and policy debate \parencite[e.g.][]{fletcher2021consumer,calvano2021market} has focused on.
However, in the more recent years, another important and potentially concerning behavior has been getting more attention: the hybrid operation of platforms.
Essentially, this means that platforms act as an intermediary (deciding the rules) and a market participant (competing with other entrants) at the same time.

While hybrid platforms are not a completely new phenomenon, they are becoming more and more common in various industries.
Perhaps the most prominent example is Amazon, which on the one hand selling its own products, and at the same time is hosting  large number of third-party sellers.
Examples abound in other industries, as well.
For example, all of the largest digital distribution platforms for computer and smartphone applications (such as Google Play, Apple App Store or Microsoft Store) sell applications products by the platform owner itself besides their third party offerings.
In the video game industry, platform owners owning their own game studios is the standard.
Many video streaming services (e.g. Netflix, Amazon Prime Video, Hulu) offer a mix of their own and licensed content.
One can also find examples of similar behavior outside the platform setting, too.
For example BMW is preparing to offer cars directly to consumers in the US, competing with its own dealership network.

As deciding the rules of the game and competing in it at the same time gives rise to some obvious concerns, policymakers have been paying a closer look at these platforms in the recent years.
Various pieces of legislation have been proposed to regulate, or even ban e-commerce platforms from selling their own products \parencite[][]{phartiyal_2019,reynolds_2022,eu-2022}.
Furthermore, one of the most prominent ongoing antitrust cases (in 2022--2023), Microsoft attempting to acquire the video game publisher Activision/Blizzard for a sum of 69 billion USD, is also a case of a platform becoming increasingly more hybrid.
This merger would not only increase market concentration on the software side of the video game industry, but might also have further implications for bargaining power between Microsoft and the other game developers.
The deal currently is investigated by multiple antitrust authorities, including the US Federal Trade Commission, the European Commission, and the UK Competition and Markets Authority \parencite{livni_merced_2023}.

In parallel to the increased regulatory interest, the academic literature has also been steadily growing on the topic of hybrid platforms.
\textcite{hagiu2022should} investigates how practices, such as self-preferencing and steering can distort competition on hybrid platforms.
They argue that when such behavior is regulated, the platform selling its own products can be beneficial for consumers.
On the contrary, \textcite[]{anderson2021hybrid} find that even in the absence of such behavior, allowing hybrid operation might have negative consequences due to platforms having an incentive to exclude competitors from the market.
\textcite[postnote]{gutierrez2021welfare} shows that general conclusions are hard to draw, and welfare effects are platform and product specific.

This paper aims to contribute to this discussion by examining an important, but so far underexplored aspect of platforms: the bargaining power disparity between them and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
In particular, I propose an analytically tractable framework which captures many important aspects of bargaining between one large and a continuum of small players, and apply it to the setting of a hybrid platform facing a continuum of potential entrants.
The results show that, in the presence of bargaining, hybrid operation can indeed have negative welfare consequences for consumers.
This result is particularly striking in this otherwise distortion-free\footnote{
    In the benchmark model without bargaining, the platform's extent of hybrid operation does not affect consumer welfare.
} setting. 
The intuition is that hybrid operation increases the platform's bargaining power due to having its own products.
This in turn leads to a higher entry fee, leading to fewer entrants, which in turn leads to lower consumer surplus.
These observations highlight a so-far overlooked aspect of hybrid platforms that policymakers should be aware of.

The closest paper to this one is \textcite{anderson2021hybrid}, which utilizes a similar model and reaches similar conclusions.
I adapt many of the main features of thir model with a couple of crucial differences.
Instead of the platform setting a percentage entry fee and being able to commit to it, I assume that (1) platforms charge a lump-sum entry fee and that (2) this entry fee is the result of a negotiation process\footnote{
    The bargaining process is modeled using a concept from cooperative game theory, namely the Shapley-value.
    This allows for a tractable analysis while at the same time capturing many of the important features of bargaining.
    Examples of this approach in the industrial organization literature include \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
}
between the platform and the entrants.
The second point is a novelty not only compared to \textcite{anderson2021hybrid}, but also to the existing literature on hybrid platforms.
Furthermore, in contrast to \textcite{hagiu2022should}, this model is free from any distortive behavior, such as self-preferencing or steering, and the negative results are purely driven by the platform's bargaining position.

This paper is also related to various other strands of the industrial organization literature.
In broad terms, it belongs to research agenda on understanding platforms and their role in the economy \parencite[e.g.][]{rochet2003platform,hagiu2004optimal,armstrong2006competition,evans2011platform,lee2014competing}.
Furthermore, it is somewhat adjacent to the literature on the importance of exclusive content \parencite[e.g.][]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}, with the difference that instead of exclusive content giving an advantage against competing platforms, in this paper, own products provide an advantage over entrants.
Somewhat more generally, many results and concepts of this paper are also applicable outside the context of (multi-sided) platforms.
Certain structures, such as vertical relationships, or even traditional retail stores share a number of features with platforms (notably, the presence of a dominant, indispensable entity), and can be modeled in much the same way.
Therefore, the methodology and results in this paper may also be of interest to researchers studying questions such as retailers having their own private labels \parencite{steiner2004nature} or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
\textcite{de2005vertical} and \textcite{montez2007downstream}, in which bargaining between the upstream and downstream firms take the center stage, are particularly closely related.

Finally, this paper also belongs to the not too large set of models using concepts from cooperative game theory in an industrial organization setting.
Such examples include the aforementioned \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
In contrast to the majority of those, which focus on games with a finite number of players, I utilize an oceanic game (a continuum of small players instead of a finite number), and demonstrate that this can considerably simplify the analysis in certain cases.
Therefore, the modeling approach presented in this paper may be of use in other settings as well.
